---
title: "Data Modification"
author: "Bill Perry"
date: "2018/03/14"
output:
  html_document:
    toc: true
    toc_float: true
---
## Two Way Anovas

###Ok so this is a completely fake dataset that I will update soon to make it more useful to illustrate Type I vs Type III errors but for now we will go with it.



###This section is for TWO WAY ANOVAs   
#### The items covered in this page are       
1. [Load libraries](#load)    
2. [Make separate Dataframes](#make)  
3. [Make one dataframe and clean up](#bind)    
4. [Get ready for ANOVA](#factors)       
5. [Boxplots](#box)    
6. [Test for homogeneity of variances](#homogeneity)   
7. [Tests for normality](#normality)     
8. [New summary statistics with RMISC](#summary_stats)  
9. [Effect plots](#plots)   
    i. [main effects](#main_effects)  
    ii. [Interaction Plots ](#interact_plots)  
10. [Finally the two way ANOVA](#two_way_aov)
    i. [AOV summary](#summary)
    ii. [Type I - WRONG if unblaanced](#type_I)
11. [Assumptions of the model](#assumptions)
12. [Post Hoc tests](#post_hoc)
    i. Main effects only
    ii. [Interaction effects ](#post_hoc_interact)

####An absolutley great coverage of this material is Dolph Schluters page at:
https://www.zoology.ubc.ca/~schluter/R/fit-model/

The key thing here is the use of the car package as it is essential for unbalanced designs and the use of Type III sum of squares otherwise Type I sum of squares are used which is rarely good. Weather to use Type II or III is a contentious issue and we will just go with Type III for all of our work

###Load the libraries and data {#load}
```{r, message=FALSE, warning=FALSE}
# Standard use packages
library(tidyverse)
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)


#specific to anova and 2 way anova
library(broom)
library(car)
library(Rmisc)
library(Hmisc)
library(lsmeans)
# library(pwr)
# library(multcompView) # interferes with DPLYR load at the very end if we need to


# so rather han load data we are going to make it up from scratch:
# set up the sample sizes for each of the fixed varaibles we have
# male OR in lakes
orml=50
# female OR in lakes
orfl=50
# male OR in streams
orms=50
# female OR in streams
orfs=50
```

###Now to make separate dataframes {#make}
```{r}
# DATASET 1 ---- ---- ---- ---- ---- ---- ---- ---- ---- ----
#make dataframe for lake male OR
# create 50 random numbes between 34 and 55

set.seed(3)
cl <- rnorm(orml, 38, 1.85)
sp <- replicate(orml, "or")
habitat <- replicate(orml, "lake")
sex <- replicate(orml, "male")
or_m_lake <- cbind(habitat, sp, sex, cl)

# DATASET 2 ---- ---- ---- ---- ---- ---- ---- ----
#make dataframe for lake female OR
# create 50 random numbes between 34 and 60
set.seed(55)
cl <- rnorm(orfl, 39, 1.75)
sp <- replicate(orfl, "or")
habitat <- replicate(orfl, "lake")
sex <- replicate(orfl, "female")
or_f_lake <- cbind(habitat, sp, sex, cl)

# DATASET 3 ---- ---- ---- ---- ---- ---- ---- ----
#make dataframe for stream male OR
# create 50 random numbes between 28 and 38
set.seed(95)
cl <- rnorm(orms, 33, 1.5)
sp <- replicate(orms, "or")
habitat <- replicate(orms, "stream")
sex <- replicate(orms, "male")
or_m_stream <- cbind(habitat, sp, sex, cl)

# DATASET 4 ---- ---- ---- ---- ---- ---- ---- ----
#make dataframe for stream female OR
# create 50 random numbes between 28 and 50
set.seed(75)
cl <- rnorm(orfs, 36, 2.5)
sp <- replicate(orfs, "or")
habitat <- replicate(orfs, "stream")
sex <- replicate(orfs, "female")
or_f_stream <- cbind(habitat, sp, sex, cl)
```



###Now to bind the dataframes together and clean up the code a bit {#bind}
```{r}
# Bind the dataframes together ----
#Rbind merges things top to bottom
or_ls.df <- as.data.frame(rbind(or_f_lake, or_m_lake, or_f_stream, or_m_stream))

# Make some new columns ----
#make a new column for some things below
or_ls.df <- unite_(or_ls.df, "sp_sex_habitat", c("sp", "sex","habitat"), sep="_", remove=FALSE)

#Get ride of all the garbage----
rm(cl, habitat, sex, sp, or_f_lake, or_m_lake, or_f_stream, or_m_stream)

#check dataframe and notice that cl is a factor
# sapply(or_ls.df, class)

# to convert to numeric takes this to make sure the number is preserved
# note this is a pain but it is how it has to be done.
or_ls.df$cl <- as.numeric(as.character(or_ls.df$cl))

# take the log of the carapace length
or_ls.df$logcl <- log(or_ls.df$cl)

head(or_ls.df)
```

###Now to look at the mechanics of a 2 WAY ANOVA - the set up {#factors}
```{r}
###NOW to two way anova ----
# need to make the categroical variable into factors 
or_ls.df$sex <- as.factor(or_ls.df$sex)
or_ls.df$sp <- as.factor(or_ls.df$sp)
or_ls.df$habitat <- as.factor(or_ls.df$habitat)
or_ls.df$sp_sex_habitat <- as.factor(or_ls.df$sp_sex_habitat)
```

###Boxplots {#box}
```{r}
# Look at the data with box plots
or_ls.df %>%
  ggplot() +
  geom_boxplot(aes(x= sp_sex_habitat, y=cl)) +
  labs(y="Carapace length")
```

###Test for homogenetiy of variances {#homogeneity} 
```{r}

# Best to use the Levenes test compared to the Bartlet test    
# uses the car package   

# http://www.cookbook-r.com/Statistical_analysis/Homogeneity_of_variance/    
 
# #Test for homogeneity of variances by groups
leveneTest(cl ~ sp_sex_habitat, data=or_ls.df)

#Compared to the Bartlet test
bartlett.test(cl ~ sp_sex_habitat, data=or_ls.df)
```

THIS IS FOR 2 Way AOV
With multiple independent variables, the interaction() function 
 must be used to collapse the IV’s into a single variable with all 
 combinations of the factors. If it is not used, then the will be the 
 wrong degrees of freedom, and the p-value will be wrong.

###Test for normality of data {#normality}

```{r}
# SHAPIRO_WILK TEST FOR NORMALITY----
# turn off scientific notaton
options(scipen = 999)
# to turn back on
#options(scipen = 0)

# Test for normality of each group and store in shapirowilktests
# This uses the broom package to get clean output of the test
or_ls.df %>% group_by(sex, habitat) %>% do(tidy(shapiro.test(.$cl)))
# they are a bit off surprisingly but go with it anyway and test residuals
# to see if the result is the same

```

### Some basic Statistics in a new way {#summary_stats}

```{r}
#Summary Stats----
# Do this to get the for N, Means Sd, Se, and CI
# need Rmisc library installed to do this
summary = summarySE(or_ls.df, 
                measurevar="cl", 
                groupvars=c("habitat","sex"))
summary
```

### Lets look at the basic plots of the effects and interactions{#main_effects}
```{r}
#Simple box plot of main effect and interaction
# this is the main effect of sit and season
boxplot(cl ~ sex,
        data = or_ls.df,
        xlab = "Sex",
        ylab = "carapace length")
```

```{r}
#Simple box plot of main effect and interaction
# this is the main effect of sit and season
boxplot(cl ~ habitat,
        data = or_ls.df,
        xlab = "Habitat",
        ylab = "carapace length")
```

```{r}
# box plots of the site and season
boxplot(cl ~ habitat:sex,
        data = or_ls.df,
        xlab = "Habitat",
        ylab = "carapace length")
```

###Interaction Plots {#interact_plots}
```{r}
# Interaction Plots----
# This will generate two plots to compare groups according to the 
# and examine interactions
# par(mfrow=c(1,3)) could use this to make multiple per page but is not great
interaction.plot(or_ls.df$habitat, or_ls.df$sex, or_ls.df$cl)
```

```{r}
interaction.plot(or_ls.df$sex,  or_ls.df$habitat, or_ls.df$cl)
```


###TWO WAY AOV MODEL----  
 This is the two way ANOVA  
What you are looking for in here is the * after the value. if that is there then
it is signficant. Below you would first look at the one where it is
site:season - because if this is signifcant this is all you will analyze in
the post anova or post F tests. If the interaction of site:season is not
significant you would then look at the individual factors of season or site
indiependently.   
   
The way to read the results below is that the interaction and sites are
not significant but season is significant and since there are only two you
really dont need to do the post F test as it will just say the same thing but
it woudl be differnet if you and more than one season for instance.   
   
    
### NOW FOR THE 2 WAY ANOVA {#two_way_aov}
```{r}
#THE TWO WAY AOV----
# Here is a two way AOV looking at site and season with the pairwise comparisons at the end
# Setting something as FACTOR makes it so that you are looking at these as categorical varaibles to compare the means of the two
# Type III is used for unbalanced designs when there are unequal numbers of samples in teh various categories or groups which is an unbalanced design


# Set it up for Type III SS ANOVA
options(contrasts = c("contr.sum", "contr.poly"))

# Fit the linear model and conduct ANOVA
model = lm(cl ~ sex*habitat, data=or_ls.df)
Anova(model, type="III")       # Use type="III" ALWAYS!!!!


```


###Now a summary of the model {#summary}
```{r}
# Are they the same? = Try running model with different data

# this produces R-Squared and overall p-value and parameter estimates
summary(model)     # Produces r-square, overall p-value, parameter estimates

```

{#type_I}
```{r}
## THIS WILL DO A TYPE I ANOVA
 anova(model)      # does the type I ANOVA 
# NOTICE THE LOWER CASE A!!!!!
 
```


### Checking assumptions of the model   {#assumptions}
#### Histogram of residuals from a linear model.    
#  The distribution of these residuals should be approximately normal.
```{r}
hist(residuals(model), 
     col="darkgray")
```


### Graphic test for normal residuals
```{r}
qqnorm(model$res)
```


### test for normality----
```{r}
shapiro.test(model$res)
```

# A plot of residuals vs. predicted values.
# The residuals should be unbiased and homoscedastic.
# For an illustration of these properties, see this diagram by
# Steve Jost at DePaul University:
#  # condor.depaul.edu/sjost/it223/documents/resid-plots.gif.
```{r}  
plot(model,1)
```

#Equality of variances with Levenes
# Check for equality of variance across the two groups. 
# If this is signficant we will have to transform the data. 
# The Levene's test uses the library(car)
```{r}
leveneTest(or_ls.df$cl ~ or_ls.df$sex * or_ls.df$habitat)
```

### POST HOC ANALYSIS {#post_hoc}
# One advantage of the using the  package for post-hoc tests is that it  
# can produce comparisons for interaction effects.
```{r}
lsmseason = pairs(lsmeans(model,"habitat"), adjust="bonferroni")

### Means sharing a letter in .group are not significantly different
# when more than one categorical varaible
# cld(lsmseason, 
#     alpha=.05, 
#     Letters=letters)

test(lsmseason)
```

```{r}
lsmseason.sex = pairs(lsmeans(model,"sex"), adjust="bonferroni")

### Means sharing a letter in .group are not significantly different
# when more than one categorical varaible
# cld(lsmseason, 
#     alpha=.05, 
#     Letters=letters)

test(lsmseason.sex)
```

### POST HOC ANALYSIS {#post_hoc_interact}
# One advantage of the using the  package for post-hoc tests is that it  
# can produce comparisons for interaction effects.
```{r}
lsminteraction = emmeans(model, 
                         pairwise ~ habitat:sex,
                         adjust="bonferroni")

lsminteraction$contrasts



# test(lsmsite)
```
















<!-- Global Site Tag (gtag.js) - Google Analytics -->
<script async src="https://www.googletagmanager.com/gtag/js?id=UA-88373117-4"></script>
<script>
  window.dataLayer = window.dataLayer || [];
  function gtag(){dataLayer.push(arguments)};
  gtag('js', new Date());

  gtag('config', 'UA-88373117-4');
</script>
