---
title: "Plotting for final graphs"
author: "Bill Perry"
date: "2019/10/26"
output:
  html_document:
    toc: true
    toc_float: true
---

The goal of this slide set is to show you how to make publication ready graphs and expand on what we have done in the basic plotting series. The goal of this is to use GGPlot and later on I will show you how to use patchwork which allows more flexibility in plot layout.    

## Load libraries
```{r, message=FALSE, warning=FALSE}

# install.packages("devtools")
# devtools::install_github("thomasp85/patchwork")

# load the libraries each time you restart R
library("readxl") # read in excel files
library("tidyverse") # dplyr and piping and ggplot etc
library("lubridate") # dates and times
library("scales") # scales on ggplot ases
library("skimr") # quick summary stats
library("janitor") # clean up excel imports
library("patchwork") # multipanel graphs
```


##Read files  
## Long fromat data    
The key to plotting in GGPlot is to use long format data. Later on we will cover how to convert between wide and long fomat. You can look at wide fromat data and see the difference.  

```{r message=TRUE, warning=TRUE}
# Long format data
south_long.df <- read_csv("data/south_lake_long.csv")
head(south_long.df)
```

### Wide format data 
This is the format you typically enter data into spreadsheet is. You can use TidyR and gather or pivot_longer to convert this to the long format.    

```{r}
south_wide.df <- read_csv("data/south_lake.csv")
head(south_wide.df)
```

# Plotting long format data
When you plot data from long format data, you need to specify a grouping statement inside the aes statement like -    
group =    This will separate data      
color =    This will map a color to each group      
linetype = This will map a linetype to each group            
fill =     This will map a fill to each group     
shape =    This will map a symbl shape to each group    

```{r}
# Plotting long format data ------
# notice that it plots all the data and is sort of a mess...
# there are no groupings of cladocerans or copepods
ggplot(south_long.df, aes(date, org_l, shape = group)) + # sometimes necessary is , group = group
  geom_point()+
  geom_line()

```

### Adding color
You can add a color = statement followed by a grouping variable and it will change colors based on the number of groups.

```{r}
# Mapping a color to data groups ----
# If you add ", color=group" inside of the aes statement it will map a color to
# each group and it is sometimes necessary to add ", group = group"
ggplot(south_long.df, aes(date, org_l, color=group)) + 
  geom_point()+
  geom_line()
```


```{r}
# Mapping shape, size, and linetype------
# this also applies to shape, linestyle and sizes (you won't want to do size)
# , shape = group
# , linetype = group
# , size = group

ggplot(south_long.df, aes(date, org_l, color=group, shape = group)) + 
  geom_point()+
  geom_line()
```

### Statistical plots
Looking at distributions of data using box and whisker plots can be very helpful this can be added as a layer to the plot    
```{r}
# Now lets look at some statistical plot
# try adding in geom_boxplot()

ggplot(south_long.df, aes(group, org_l, color=group)) + 
  geom_boxplot() +
  geom_point() 
```

### Dodging points
Notice above that the pints are overlapping and adding in a position statement to the geom_point statement can fix that
```{r}
# to fix overlying points
ggplot(south_long.df, aes(group, org_l, color=group)) + 
  geom_boxplot() +
  geom_point(position= position_jitterdodge(jitter.width = 0.3))

```


### Adding custom colors, shapes, fills
Once you have mapped a color, fill, or shape to your groups you can add in a scale_color_manual, scale_fill_manual, or scale_shape_manual statement to adjust these proporties.
```{r}
# Final publication quality graph-----
# now we can add axes labels and custom colors
ggplot(south_long.df, aes(date, org_l, color=group)) + # sometimes necessary is , group = group
  geom_point()+
  geom_line()  +
  scale_color_manual(name = "Group", 
                     values = c("blue", "red"),
                     labels = c("Cladoceran", "Copepod")) +
  theme(axis.text.x = element_text(size=12, face="bold", angle=45, hjust=1))
```


### Adding custom scales on the axes
You can add in custom scales to the differnet axes. These can be
scale_x(or y)_date    
scale_x(or y)_datetime    
scale_x(or y)_continuous     
scale_x(or y)_log    
etc.....  

```{r}
# Final publication quality graph-----
# now we can add axes labels and custom colors
ggplot(south_long.df, aes(date, org_l, color=group)) + # sometimes necessary is , group = group
  geom_point()+
  geom_line() +
  labs(x = "Date", y = expression(bold("Animals (No. L"^-1*")"))) +
  scale_x_date(date_breaks = "3 month",
               limits = as_date(c('1994-06-01', '2006-12-31')),
               labels=date_format("%Y-%m-%d"), expand=c(0,0))   +
  scale_color_manual(name = "Group", 
                     values = c("blue", "red"),
                     labels = c("Cladoceran", "Copepod")) +
  theme(axis.text.x = element_text(size=12, face="bold", angle=45, hjust=1))
```
      
         
         

You can also change the data breaks in a simple manner 
```{r}
# Note here the labels on the X Axis are too frequent - try changing this below....
# try 6 months, 12 months, 1 years, 52 weeks 
ggplot(south_long.df, aes(date, org_l, color=group)) + # sometimes necessary is , group = group
  geom_point()+
  geom_line() +
  labs(x = "Date", y = expression(bold("Animals (No. L"^-1*")"))) +
  scale_x_date(date_breaks = "52 weeks",
               limits = as_date(c('1994-06-01', '2006-12-31')),
               labels=date_format("%Y-%m-%d"), expand=c(0,0)) +
  scale_color_manual(name = "Group", 
                     values = c("blue", "red"),
                     labels = c("Cladoceran", "Copepod"))  +
  theme(axis.text.x = element_text(size=12, face="bold", angle=45, hjust=1))

```         


### Adding in custom axes labels
You can take axes labels to the next step with the expression statment.
```{r}
# Final publication quality graph-----
# now we can add axes labels and custom colors
ggplot(south_long.df, aes(date, org_l, color=group)) + # sometimes necessary is , group = group
  geom_point()+
  geom_line() +
  labs(x = "Date", y = expression(bold("Animals (No. L"^-1*")"))) +
  scale_x_date(date_breaks = "3 month",
               limits = as_date(c('1994-06-01', '2006-12-31')),
               labels=date_format("%Y-%m-%d"), expand=c(0,0))  +
  scale_color_manual(name = "Group", 
                     values = c("blue", "red"),
                     labels = c("Cladoceran", "Copepod")) +
  theme(axis.text.x = element_text(size=12, face="bold", angle=45, hjust=1)) 
```




## Facetting graphs      
You can also break up data by facetting the data.   
The use of facet_wrap and facet_grid only differ slightly   
### Facet_wrap
```{r}
# Facetting graphs -----
# So lets begin to break it up a bit more... this is a busy graph... what
# if we wanted to see two separate graphs?

# Facet wrap ---- 
# sort of flows graphs around by a variable
ggplot(south_long.df, aes(date, org_l, color=group)) + # sometimes necessary is , group = group
  geom_point()+
  geom_line() +
  labs(x = "Date", y = expression(bold("Animals (No. L"^-1*")"))) +
  scale_x_date(date_breaks = "12 month",
               limits = as_date(c('1994-06-01', '2006-12-31')),
               labels=date_format("%Y-%m-%d"), expand=c(0,0))  +
  theme(axis.text.x = element_text(size=12, face="bold", angle=45, hjust=1)) +
  scale_color_manual(name = "Group", 
                     values = c("blue", "red"),
                     labels = c("Cladoceran", "Copepod")) +
  facet_wrap(~group)

```

### Facet_grid
```{r}
# Facet_grid -----
# lets you make a grid of one or two variables in a grid

ggplot(south_long.df, aes(date, org_l, color=group)) + # sometimes necessary is , group = group
  geom_point()+
  geom_line() +
  labs(x = "Date", y = expression(bold("Animals (No. L"^-1*")"))) +
  scale_x_date(date_breaks = "12 month",
               limits = as_date(c('1994-06-01', '2006-12-31')),
               labels=date_format("%Y-%m-%d"), expand=c(0,0))  +
  theme(axis.text.x = element_text(size=12, face="bold", angle=45, hjust=1)) +
  scale_color_manual(name = "Group", 
                     values = c("blue", "red"),
                     labels = c("Cladoceran", "Copepod")) +
  facet_grid(.~group) # rows by columns
  

```

### Facet_grid vertical
```{r}
# Facet_grid vertical ----
ggplot(south_long.df, aes(date, org_l, color=group)) + # sometimes necessary is , group = group
  geom_point()+
  geom_line() +
  labs(x = "Date", y = expression(bold("Animals (No. L"^-1*")"))) +
  scale_x_date(date_breaks = "12 month",
               limits = as_date(c('1994-06-01', '2006-12-31')),
               labels=date_format("%Y-%m-%d"), expand=c(0,0))  +
  theme(axis.text.x = element_text(size=12, face="bold", angle=45, hjust=1)) +
  scale_color_manual(name = "Group", 
                     values = c("blue", "red"),
                     labels = c("Cladoceran", "Copepod")) +
  facet_grid(group~.) # rows by columns

```

### Labeller for facets
You can also do custom lables and titles for each facet using labeller.
```{r}
# Relabelling facet titles -----
# now if we wanted to relabel the graphs we can do that as well

ggplot(south_long.df, aes(date, org_l, color=group)) + # sometimes necessary is , group = group
  geom_point()+
  geom_line() +
  labs(x = "Date", y = expression(bold("Animals (No. L"^-1*")"))) +
  scale_x_date(date_breaks = "12 month",
               limits = as_date(c('1994-06-01', '2006-12-31')),
               labels=date_format("%Y-%m-%d"), expand=c(0,0))  +
  theme(axis.text.x = element_text(size=12, face="bold", angle=45, hjust=1),
        strip.background = element_rect(fill = NA),
        strip.text = element_text(size = 14, face = "bold", hjust = 0)) +
  scale_color_manual(name = "Group", 
                     values = c("blue", "red"),
                     labels = c("Cladoceran", "Copepod")) +
  facet_grid(.~group,
             labeller = labeller(group = c("cladoceran" = "a) Cladoceran",
                                             "copepod" = "b) Copepod"))) # rows by columns

```









<!-- Global site tag (gtag.js) - Google Analytics -->
<script async src="https://www.googletagmanager.com/gtag/js?id=UA-88373117-6"></script>
<script>
  window.dataLayer = window.dataLayer || [];
  function gtag(){dataLayer.push(arguments);}
  gtag('js', new Date());

  gtag('config', 'UA-88373117-6');
</script>
