---
title: "Libraries"
author: "Bill Perry"
date: "2019/06/10"
output:
  # css: stats.css
  html_document:
   toc: TRUE
   toc_float: TRUE
   number_sections:  FALSE
   
---

# Libraries     
These add functionality to base R. This is similiar to adding apps on your phone.

## These are in my mind the essential libraries for R
```{r Essential_Libraries, eval=FALSE}
# # Libraries I use on a regular basis-----
install.packages("devtools") # install new things
install.packages("tidyverse") # dplyr and piping and ggplot etc
install.packages("lubridate") # dates and times
install.packages("scales") # scales on ggplot ases
install.packages("readxl") # read in excel files
install.packages("skimr") # quick summary stats
install.packages("hms") # useful for time series
install.packages("janitor") # clean up excel imports

devtools::install_github("thomasp85/patchwork") # multiple plots
devtools::install_github("calligross/ggthemeassist") # ggplot adjust
```

## These libraries are useful for general statistics and used quite often    
```{r General_Statistics, eval=FALSE}
install.packages("car") # stats and ANOVA - essential
install.packages("emmeans") # estimated marginal means for unbalanced designs
install.packages("multcomView") # paired comparisons
install.packages("Rmisc") # stats
install.packages("Hmisc") # stats

install.packages("broom") # output models cleanly
install.packages("GGally") # special ggplot graphs
install.packages("corrplot") # correlation plot matricies
install.packages("gvlma") # regressions and assumptions
```

## These are specialized libraries that I often use    
```{r special_packages, eval=FALSE}

install.packages("akima") # inputing new values
install.packages("survminer") # survival analysis
install.packages("survival") # survival analysis
install.packages("styler") # style your code - nice
install.packages("plotly") # cool ggplot things
install.packages("colorRamps")


install.packages("pwr") # power analysis

install.packages("vegan") # PCA and other sorts of multidimensional
install.packages("factoextra")

# mixed models
install.packages("lmerTest")
install.packages("blme") 


```


## Finally here are some lesser used pacakges 

```{r infrequent_packages, eval=FALSE}
# these are special that you likely will not use 
# devtools::install_github("seasmith/AlignAssign")
# install.packages("data.table")
# install.packages("RMySQL") # access MySQLserver
# install.packages("xts")  # not used
# install.packages("rLakeAnalyzer") # used for the lake analyzer scripts
# install.packages("LakeMetabolizer")
# install.packages("zoo")

```

## To get the libraires running - some need to only be run one time
```{r libraries_once, eval=FALSE}
# Run Once and in Addins ----
library(styler)
library(ggThemeAssist)

```

## Other libraries I use each and ever time
```{r typical_libraries}
# Load in all scripts ---
library(tidyverse)	
library(lubridate)	
library(scales)	
library(readxl)
library(skimr)
library(janitor)
library(patchwork)

```

<!-- Global Site Tag (gtag.js) - Google Analytics -->
<script async src="https://www.googletagmanager.com/gtag/js?id=UA-88373117-4"></script>
<script>
  window.dataLayer = window.dataLayer || [];
  function gtag(){dataLayer.push(arguments)};
  gtag('js', new Date());

  gtag('config', 'UA-88373117-4');
</script>
