---
title: "Intermediate GGPlot Stats"
author: "Bill Perry"
date: "2018/03/14"
output:
  html_document:
    toc: true
    toc_float: true
---

##Load libraries
##Read files
```{r, message=FALSE, warning=FALSE}
# One new package for summary stats
#install.packages("psych")

# load the libraries each time you restart R
library(tidyverse)
library(lubridate)
library(scales)
library(skimr)


# read in the file
iris.df <- read_csv("data/iris.csv")

glimpse(iris.df)
```

##Simple histograms

```{r}
#Hitograms with GGPLOT
ggplot(data=iris.df, aes(iris.df$Sepal.Length)) + 
  geom_histogram() +
  labs (title="Sepal Length", x="Sepal Length", y="Count")
```


##Histogram binwiths
```{r}
#Hitograms with GGPLOT
ggplot(data=iris.df, aes(iris.df$Sepal.Length)) + 
  geom_histogram(col="blue", fill="blue", alpha=.7, binwidth = 1) +
  labs (title="Sepal Length", x="Sepal Length", y="Count")
```

##Fccetted histograms
```{r}
###What if you want to look at all of the species separately
ggplot(data=iris.df, aes(iris.df$Sepal.Length)) + 
  geom_histogram(binwidth = .5,  col="blue", fill="blue", alpha=.7) +
  labs (title="Sepal Length", x="Sepal Length", y="Count") + 
  facet_wrap(~Species, scales = "free") # allows the scales to be free and you can comment this out to see what happens

```

##Other visualizations
###Dot plots


```{r}
ggplot(data=iris.df, aes(x= Species, y=Sepal.Length)) + 
  geom_dotplot(binwidth=0.1, col="blue", fill="blue", alpha=.7, stackdir = "center", binaxis = "y") +
  labs (title="Sepal Length", x="Sepal Length", y="Count") 
```

###Box and Whisker
```{r}
# Box and Whisker plots with limits ####
ggplot(iris.df, aes(x = Species, y = Sepal.Length))+
  geom_boxplot() 
```

###Vioin plots
```{r}
ggplot(data=iris.df, aes(x= Species, y=Sepal.Length)) + 
  geom_violin(col="blue", fill="blue", alpha=.7, stackdir = "center", binaxis = "y") +
  labs (title="Sepal Length")+
  labs (x="Sepal Length", y="Count") 
```

###Saving plots 
```{r}

violin.plot <- ggplot(data=iris.df, aes(x= Species, y=Sepal.Length)) + 
  geom_violin(col="blue", fill="blue", alpha=.7) +
  labs (title="Sepal Length")+
  labs (x="Sepal Length", y="Count") 
violin.plot

#Save the file with sizes
ggsave("figures/violin.plot.tiff",plot=violin.plot, width=6, height=6, dpi = 300)



```

##Statistical plots     
So now lets try to do a few plots that are useful for publication. the first will be a mean and standard error plot and we will try this with multiple data in wide format

#Plotting multiple variables - wide format
First lets just plot Species versus Sepal Length.
```{r}
ggplot(iris.df, aes(x=Species)) +
  geom_point(aes(y=Sepal.Length))
```

Now lets add Sepal.Width
```{r}
ggplot(iris.df, aes(x=Species)) +
  geom_point(aes(y=Sepal.Length), color="blue") +
  geom_point(aes(y=Sepal.Width), color="red")
```

##Means
Now what if you wanted only to show the means
```{r}
ggplot(iris.df, aes(x=Species))+
  stat_summary(aes(y=Sepal.Length), fun.y=mean, geom='point', shape=23, size=3, color="red", fill="red") 
```

Add in the second mean
Now what if you wanted only to show the means
```{r}
ggplot(iris.df, aes(x=Species))+
  stat_summary(aes(y=Sepal.Length), fun.y=mean, geom='point', shape=23, size=3, color="red", fill="red") +
    stat_summary(aes(y=Sepal.Width), fun.y=mean, geom='point', shape=23, size=3, color="red", fill="blue") 
```

Now your boss of course wants standard error bars on each mean... no problem

```{r}
ggplot(iris.df, aes(x=Species))+
  stat_summary(aes(y=Sepal.Length), fun.y=mean, geom='point', shape=23, size=3, color="red", fill="red") +
    stat_summary(aes(y=Sepal.Width), fun.y=mean, geom='point', shape=23, size=3, color="blue", fill="blue") +
   stat_summary(aes(y=Sepal.Length), fun.data = mean_se, geom = "errorbar", color="red", width=.1) +
    stat_summary(aes(y=Sepal.Width),fun.data = mean_se, geom = "errorbar", color="blue", width=.1) 

```

you could also do standard deviation
```{r}
ggplot(iris.df, aes(x=Species))+
  stat_summary(aes(y=Sepal.Length), fun.y=mean, geom='point', shape=23, size=3, color="red", fill="red") +
    stat_summary(aes(y=Sepal.Width), fun.y=mean, geom='point', shape=23, size=3, color="blue", fill="blue") +
   stat_summary(aes(y=Sepal.Length), fun.data = mean_sdl, geom = "errorbar", color="red", width=.1) +
    stat_summary(aes(y=Sepal.Width),fun.data = mean_sdl, geom = "errorbar", color="blue", width=.1) 

```

###Position_dodge - move points
Now you boss says great but the error bars overlap... can you fix this?

```{r}
ggplot(iris.df, aes(x=Species))+
  stat_summary(aes(y=Sepal.Length), fun.y=mean, geom='point', shape=23, size=3, color="red", 
               fill="red", position=position_nudge(x = 0.12, y = 0)) +
    stat_summary(aes(y=Sepal.Width), fun.y=mean, geom='point', shape=23, size=3, color="blue", 
                 fill="blue") +
   stat_summary(aes(y=Sepal.Length), fun.data = mean_sdl, geom = "errorbar", color="red", width=0.1,
                position=position_nudge(x = 0.12, y = 0)) +
    stat_summary(aes(y=Sepal.Width),fun.data = mean_sdl, geom = "errorbar", color="blue", width=0.1) 

```








<script>
  (function(i,s,o,g,r,a,m){i['GoogleAnalyticsObject']=r;i[r]=i[r]||function(){
  (i[r].q=i[r].q||[]).push(arguments)},i[r].l=1*new Date();a=s.createElement(o),
  m=s.getElementsByTagName(o)[0];a.async=1;a.src=g;m.parentNode.insertBefore(a,m)
  })(window,document,'script','https://www.google-analytics.com/analytics.js','ga');

  ga('create', 'UA-88373117-2', 'auto');
  ga('send', 'pageview');

</script>