---
title: "Means and standard error plotting"
author: "Bill Perry"
date: "2019/10/26"
output:
  html_document:
    theme: cosmo
    toc: TRUE
    toc_float: TRUE
---

## Load libraries

```{r, message=FALSE, warning=FALSE}

# install.packages("devtools")
# devtools::install_github("thomasp85/patchwork")

# load the libraries each time you restart R
library("readxl") # read in excel files
library("tidyverse") # dplyr and piping and ggplot etc
library("lubridate") # dates and times
library("scales") # scales on ggplot ases
library("skimr") # quick summary stats
library("janitor") # clean up excel imports
library("patchwork") # multipanel graphs
library(skimr) # great way to do summary stats
``` 


##Read files

```{r}
# lets read in a new file to add some complexity for fun
mm.df <- read_csv("data/mms.csv")


```

# Summary Stats
Lets look at a few ways to get summary statistics
The first is the simplist and uses base R
```{r}
summary(mm.df)
```

A better way is using Skimr
```{r}
mm.df %>% 
  skim()
```

The cool part of skimr is that you can do groups
```{r}
mm.df %>% 
  group_by(center) %>% 
  skim()
```

Finally you can get a summary a differnt way but is a bit longer
```{r}
mm.df %>% 
  group_by(center, color) %>% 
  summarize(mean_diamter = mean(diameter, na.rm=TRUE),
            mean_mass = mean(mass, na.rm=TRUE))
```

# Graphing mena and SE

These are all well and good but looking at a graph is cool.

```{r}
# now for the plot
ggplot(mm.df, aes(color, diameter, color=color)) + 
  stat_summary(fun = mean, na.rm = TRUE,
               geom = "point",
               size = 3) + 
  stat_summary(fun.data = mean_se, na.rm = TRUE,
               geom = "errorbar",
               width = 0.2) +
  labs(x = "color", y = "Diamter (units") 
```

We can also add in shape as a grouping varaible for the center of the m&m's

```{r}
# now for the plot
ggplot(mm.df, aes(color, diameter, group=center, shape=center, color=color)) + 
  stat_summary(fun = mean, na.rm = TRUE,
               geom = "point",
               size = 3) + 
  stat_summary(fun.data = mean_se, na.rm = TRUE,
               geom = "errorbar",
               width = 0.2) +
  labs(x = "color", y = "Diamter (units)") 
```






<!-- Global site tag (gtag.js) - Google Analytics -->
<script async src="https://www.googletagmanager.com/gtag/js?id=UA-88373117-6"></script>
<script>
  window.dataLayer = window.dataLayer || [];
  function gtag(){dataLayer.push(arguments);}
  gtag('js', new Date());

  gtag('config', 'UA-88373117-6');
</script>
