---
title: "Intermediate GGPlot Stats"
author: "Bill Perry"
date: "2021/06/15"
output:
  html_document:
    theme: cosmo    
    toc: true
    toc_float: true
---

## Load libraries

```{r, message=FALSE, warning=FALSE}

# install.packages("devtools")
# devtools::install_github("thomasp85/patchwork")

# load the libraries each time you restart R
library("readxl") # read in excel files
library("tidyverse") # dplyr and piping and ggplot etc
library("lubridate") # dates and times
library("scales") # scales on ggplot ases
library("skimr") # quick summary stats
library("janitor") # clean up excel imports
library("patchwork") # multipanel graphs
```


##Read files
```{r}
# So now we have seen how to look at the data
# What if we wanted to modify the data in terms of columns or rows

# Making graphs this way can get a bit cumbersome as you might imagine. 
# This is because the data is in what we call wide format 
# The long format is the format often used for Anovas and other stats
# We will go over how to do this later but for now lets just look at the file

mm.df <- read_csv("data/mms.csv")
```

```{r}
# Plotting long format data ------
# notice that it plots all the data and is sort of a mess...
# there are no groupings of cladocerans or copepods

ggplot(mm.df, aes(color, mass)) + # sometimes necessary is , group = group
  geom_point()

```


```{r}
# Mapping a color to data groups ----
# If you add ", color=group" inside of the aes statement it will map a color to
# each group and it is sometimes necessary to add ", group = group"
ggplot(mm.df, aes(color, mass, color=color)) + 
  geom_point()
```

# Adding grouping aestethics
We know shape is in there so we should add that
```{r}
ggplot(mm.df, aes(color, mass, color=color, shape = center)) + 
  geom_point()
```
# Dodging points
Now lest try to dodge the points

```{r}
ggplot(mm.df, aes(color, mass, color=color, shape = center)) + 
  geom_point(position= position_jitterdodge(jitter.width = 0.4))
```


```{r}
# Now lets look at some statistical plot
# try adding in geom_boxplot()

ggplot(mm.df, aes(color, mass, color=color, shape = center)) + 
  geom_point(position= position_jitterdodge(jitter.width = 0.4)) 

```

```{r}
# Final publication quality graph-----
# now we can add axes labels and custom colors
ggplot(mm.df, aes(color, mass, color=color, shape = center)) + 
  geom_boxplot() +
  geom_point(position= position_jitterdodge(jitter.width = 0.4)) +
  labs(x = "color", y = "Mass") +
  scale_color_manual(name = "Color", 
                     values = c("blue", "brown", "green", "orange", "red", "yellow"),
                     labels = c("Blue", "Brown", "Green", "Orange", "Red", "Yellow"))
```
# Facetting graphs    
what if we wanted to break up this graph

```{r}
ggplot(mm.df, aes(color, mass, color=color, shape = center)) + 
  geom_boxplot() +
  geom_point(position= position_jitterdodge(jitter.width = 0.4)) +
  labs(x = "color", y = "Mass") +
  scale_color_manual(name = "Color", 
                     values = c("blue", "brown", "green", "orange", "red", "yellow"),
                     labels = c("Blue", "Brown", "Green", "Orange", "Red", "Yellow"))+
  facet_wrap(~center)

```


```{r}
# Facet_grid -----
# lets you make a grid of one or two variables in a grid

ggplot(mm.df, aes(color, mass, color=color, shape = center)) + 
  geom_boxplot() +
  geom_point(position= position_jitterdodge(jitter.width = 0.4)) +
  labs(x = "color", y = "Mass") +
  scale_color_manual(name = "Color", 
                     values = c("blue", "brown", "green", "orange", "red", "yellow"),
                     labels = c("Blue", "Brown", "Green", "Orange", "Red", "Yellow"))+
  facet_grid(center~., scales="free_y")

```




<script>
  (function(i,s,o,g,r,a,m){i['GoogleAnalyticsObject']=r;i[r]=i[r]||function(){
  (i[r].q=i[r].q||[]).push(arguments)},i[r].l=1*new Date();a=s.createElement(o),
  m=s.getElementsByTagName(o)[0];a.async=1;a.src=g;m.parentNode.insertBefore(a,m)
  })(window,document,'script','https://www.google-analytics.com/analytics.js','ga');

  ga('create', 'UA-88373117-2', 'auto');
  ga('send', 'pageview');

</script>