---
title: "T Test and One Way ANOVA"
author: "Bill Perry"
date: "2018/04/30"
output:
  html_document:
    toc: true
    toc_float: true
---
##Load libraries
We will read in the main files and load the libraries as we have worked with so far.   

```{r, message=FALSE, warning=FALSE}
# #Install Packages ----
# install.packages("tidyverse")
# install.packages("lubridate")
# install.packages("scales")
# install.packages("readxl")
# install.packages("survminer")
# install.packages("survival")
# install.packages("patchwork")
# install.packages("broom")
# ANOVA specific
# install.packages("car")
# install.packages("emmeans")
# install.packages("multcompView")


#Load libraries ----
library(tidyverse)
library(lubridate)
library(scales)
library(readxl)
library(skimr)
library(broom)
library(janitor)
# library(zoo)
library(patchwork)
library(car)
library(emmeans)
library(multcompView)

```


## Read in the file
```{r}
# read in the file
gene_exp.df <- read_excel("data/Example data.XLSX", sheet="Gene expression analysis") %>%
  clean_names() %>%
  remove_empty(c("rows", "cols")) 

glimpse(gene_exp.df)
```


##Summary Statistics for the better look
```{r}
# the data you want to look at

skim(gene_exp.df)

```

##Look at the data
We could do this in the wide format but it is a lot easier in long format
## Wide to long format
```{r}
# this will add an index to the dataframe so you know what individual is which
gene_exp_long.df <- gene_exp.df %>%
  gather(
    treatment, # this will take all the row headings into a column
    expression, # this will convert all the measures into a column called expression
    -experiment, # the - sign tells tidyr not to move those columns
    -individual
  ) 

glimpse(gene_exp_long.df)

# we could spread it back out to what it was
gene_exp_wide.df <- gene_exp_long.df %>%
  spread(treatment, expression)


```


## Look at data
```{r}
glimpse(gene_exp_long.df)
```

##Convert to factors
So to do this the experiment and individuals are numeric and need to be converted to categories or factors...

## Wide data to factors

```{r}
# convert things to factors or do calculations
gene_exp.df <- gene_exp.df %>%
  mutate(
    experiment = as.factor(experiment),
    individual = as.factor(individual)
  )

# now look at it
skim(gene_exp.df)

```


## Long data to factors

```{r}
#need to make treatment a factor
gene_exp_long.df <- gene_exp_long.df %>%
mutate(
  experiment = as.factor(experiment),
    individual = as.factor(individual),
    treatment = as.factor(treatment)
  )

# now look at it
skim(gene_exp_long.df)




```



##Now to graph the long format data
```{r}
# Note the new format allows us to make coding a lot faster
gene_exp_long.df %>% 
  group_by(treatment)  %>%
  ggplot(aes(treatment, expression)) +
  geom_boxplot()
```








## T-Tests are a great way to compare two means. But they can be a bit limited.

There are several types of T tests and there are a few assumptions
Normality
Heterogeneity of Variances
and a few others we will talk about....



###Test for normality of each group
To do this in wide format it is straight forward  
Here these should be non-signficant so they are not deviating from a normal distribution.   

```{r}
# Test for normality of each group wide format
shapiro.test(gene_exp.df$control)
shapiro.test(gene_exp.df$treat1)
shapiro.test(gene_exp.df$treat2)
shapiro.test(gene_exp.df$treat3)


```

###Test for homogeneity of variances  
Best to use the Levenes test compared to the Bartlet test  
uses the car package   
http://www.cookbook-r.com/Statistical_analysis/Homogeneity_of_variance/   
```{r}
#Test for homogeneity of variances

leveneTest(expression ~ treatment, data=gene_exp_long.df)

#Compared to the Bartlet test
bartlett.test(expression ~ treatment, data=gene_exp_long.df)

#With multiple independent variables, the interaction() function must be used to collapse the IV’s into a single variable with all combinations of the factors. If it is not used, then the will be the wrong degrees of freedom, and the p-value will be wrong.

# bartlett.test(extra ~ interaction(group,some other independent factor), data=sleep.df)


```

 
  
###Paired T-Tests wide format
```{r}
# Paired T-Test - Welch 2 sample -----
t.test(gene_exp.df$control, gene_exp.df$treat1, paired=TRUE) # Note the 1 and 2 are in quotes so R knows its the column
```


###Paired T-Tests long format
```{r}
# Paired T-Test - Welch 2 sample -----
# Welch t-test
# t.test(variable ~ group, dataframe)

# need to make a dataframe with only 2 groups
cont_t1.df <- gene_exp_long.df %>%
  filter(treatment == "control" | treatment == "treat1")


t.test(expression ~ treatment, cont_t1.df, paired=TRUE) # Note the 1 and 2 are in quotes so R knows its the column
```


###Two Sample T-Test - in wide format
By default, t.test does not assume equal variances
```{r}
# Welch t-test
t.test(gene_exp.df$control, gene_exp.df$treat1, var.equal=FALSE)
```


###Two Sample T-Test - in long format
By default, t.test does not assume equal variances
```{r}
# Welch t-test
t.test(expression ~ treatment, cont_t1.df, var.equal=FALSE)
```



## One way Anova

## Great Reading
A Great coverage of this material is Dolph Schluters page
https://www.zoology.ubc.ca/~schluter/R/fit-model/

The key thing here is the use of the car package as it is essential for unbalanced designs and the use of Type III sum of squares otherwise Type I sum of squares are used which is rarely good. Weather to use Type II or III is a contentious issue and we will just go with Type III for all of our work


## Bartlets test for homogeneity of varaince
```{r}
#Compared variances uisng  Bartlet test
# significant means they differ in varainces but this is close
bartlett.test(expression ~ treatment, data=gene_exp_long.df)
```


## ANOVA
Now we can run an ANOVA as long as the categories we are testing are factors. When doing this we will test to see if any of the means are different but will not be able to tell what is different yet. Thatis the next step

```{r}
# ANOVA - ONE WAY
# you can do an anova as an anova and not the linear model
#Run the anova and store it in the model in Values
expression.model.aov = aov(expression ~ treatment, data=gene_exp_long.df)

#Obtain the anova table
anova(expression.model.aov)

```

## Save the model for use in word
```{r}

# save model to a text file for excel or whatever
tidy(expression.model.aov)

#You can copy this out or save it as an object and then save it as a csv file
# save the model
# tidy_expression.model.aov <- tidy(expression.model.aov)
# write_csv(expression.model.aov, "tidy_anova_expression.csv")

```


## Plot Residuals

```{r}
# Plot residuals
#Base R plots
plot(fitted(expression.model.aov), residuals(expression.model.aov))
```


## Histogram of residuals
```{r}
#Histogram of residuals
hist(residuals(expression.model.aov), 
     col="darkgray")
```

## Check for Normality
```{r}
# check for normally distributed data
qqnorm(expression.model.aov$res)
```

## Statistical Test of Normality
```{r}
#Test for normality of residuals
shapiro.test(expression.model.aov$res)
```

## Post F tests of an ANOVA

```{r}
# Post F tests
# Comparisons of species
lsm = emmeans(expression.model.aov, 
              "treatment",
              adjust="bonferroni")

### Means sharing a letter in .group are not significantly different
#Note that this requires multcompView
multcomp::cld(lsm,
    alpha=.05,
    Letters=letters)
```

```{r}
# Now you have a statistical test of how the means compare
gene_exp_long.df %>% 
  group_by(treatment)  %>%
  ggplot(aes(treatment, expression)) +
  geom_boxplot()
```

```{r}
# what if you wanted a bar plot
ggplot(gene_exp_long.df, aes(x= treatment))+
  stat_summary(aes(y=expression), fun.y=mean, geom='bar', color="black", 
               fill="blue") +
  stat_summary(aes(y=expression), fun.data = mean_se, geom = "errorbar", color="black", width=0.2)
```



   

<!-- Global Site Tag (gtag.js) - Google Analytics -->
<script async src="https://www.googletagmanager.com/gtag/js?id=UA-88373117-4"></script>
<script>
  window.dataLayer = window.dataLayer || [];
  function gtag(){dataLayer.push(arguments)};
  gtag('js', new Date());

  gtag('config', 'UA-88373117-4');
</script>
