---
title: "Converting between logn and wide and back again"
author: "Bill Perry"
date: "2019/10/26"
output:
  html_document:
    toc: true
    toc_float: true
---

# Load libraries

```{r}
# Load Libraries ----
# this is done each time you run a script
library("readxl") # read in excel files
library("tidyverse") # dplyr and piping and ggplot etc
library("lubridate") # dates and times
library("scales") # scales on ggplot ases
library("skimr") # quick summary stats
library("janitor") # clean up excel imports
library("patchwork") # multipanel graphs
```

# Read in files

```{r}
# So now we have seen how to look at the data
# What if we wanted to modify the data in terms of overall structure

# If you were typing in data this might be how it looks
# Read in wide dataframe ----
lakes.df <- read_csv("data/reduced_lake_long_genus_species_wide.csv")
head(lakes.df)
```

# Gather to long format      
How to convert a wide dataframe to a long data frame. This is the older approach which has since been replaced by pivot_longer. Gather is still something I use a lot.    
```{r}
# Convert wide to long format ----
# this is an older method that is simple
lakes_long.df <- lakes.df %>% 
  gather(genus_species, # this will make a new column group with the column names 
         org_l,  # this will make a column of data for the counts per liter
         -lake_name, -date, -permanent_id, -year # "-" means leave alone
  )

```


# Pivot_longer to long format     
This is the same as above and does the same thing.  
```{r}
# this is the newer way that might be better
lakes_long.df <- lakes.df %>%
  pivot_longer(
    -c(lake_name, date, permanent_id, year),
    names_to = "genus_species", 
    values_to = "org_l")
```

# Spread
This spreads a long dataframe into a wide fomat dataframe.        
```{r}
# Long to Wide format ----
# the older method 
lakes_wide.df <- lakes_long.df %>%
  spread(
    genus_species, org_l
  )
```


# Pivot_wider     
This is the same as spread using a slightly newer terminology      
```{r}
# now the newer method
lakes_wide.df <- lakes_long.df %>%
  pivot_wider(
    id_cols = c( "lake_name", "date" ),
    names_from = genus_species,
    values_from = org_l)

```


# Read in long format file   
Lets read in a new version of the file and do some more work    
```{r}
# now the new version of modification can also do cool stuff like 
# summarizing data 
# lets read in a new file to add some complexity for fun
lakes.df <- read_csv("data/reduced_lake_long_genus_species.csv")

```

Here we remove the genus species column in preparation to summarize the data
```{r}
# lets simplify the data
lakes.df <- lakes.df %>% select(-genus_species)
```


# Pivot to summarize data
We can also use the pivot_wider or pivot_longer to summarize data 

```{r}
# make wide with column headers of group and summarize by lake and date
group_wide.df <- lakes.df %>%
  pivot_wider(names_from = group, 
              values_from = org_l,
              values_fn = list(org_l = mean, na.rm=TRUE))
```






<!-- Global site tag (gtag.js) - Google Analytics -->
<script async src="https://www.googletagmanager.com/gtag/js?id=UA-88373117-6"></script>
<script>
  window.dataLayer = window.dataLayer || [];
  function gtag(){dataLayer.push(arguments);}
  gtag('js', new Date());

  gtag('config', 'UA-88373117-6');
</script>
